The evolution of gene expression in the term placenta of viviparous mammals

### 2016-04-18 manuscript/manuscript.Rnw
add Derek suggested changes

  \parttitle{Conclusions} A diverse array of genes are differentially
  expressed and are likely to play a role in the anatomical
  variability of the placenta in the placentas of mammalian lineages.
  However, the expression of a subset of
  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
  genes is essential for the proper functioning of the placenta in all
  species analyzed.

@@

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form
\cite{Mossman.1937:comparativemorphogenesisofthe,Moffett.ea2006:Immunologyofplacentationin}. Species differ in
the shape of attachment (i.e., discoid, diffuse, cotyledonary,
zonary), degree of invasiveness of fetal tissues into maternal tissues
(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
interdigitation of maternal-fetal tissue (i.e., villous, folded,
trabecular, lamellar, labrynthine)
\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
mammals differ drastically in the number and size of their offspring,
as well as the length of gestation, traits often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(\cref{tab:gestational_periods}).

The placenta arose at some point before the origin of extant mammals,
a group that includes marsupials, monotremes, and eutherian or
“placental mammals”. The term “placental mammals” is misleading, as
extant yinotherians (such as \textit{Ornithorhynchus anatinus}, the
platypus) possess a rudimentary placenta, as do marsupials, although
the direct apposition to maternal tissues lasts for a proportionally
shorter time during gestation
\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
Marsupials also differ from eutherian mammals in that most possess
only a choriovitelline placenta, produced from the fusion of the
chorion and yolk sac, during most of gestation. Although some
eutherian mammals may possess a choriovitelline placenta at some point
in development, the major site of nutrient exchange in placentals
occurs via the chorioallantoic placenta, produced from the fusion of
the chorion and allantois
\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
the first placenta was more like that of marsupials or eutherians;
however, using phylogenetic methods, the ancestral eutherian was
reconstructed as possessing a relatively invasive hemochorial
placenta, as in humans and mouse
\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
in \cow, genes related to the thymus-specific serine protease
\textit{PRSS16} in \textit{Loxodona africana} (in this work),
 and the expansion of the growth hormone
and chorionic gonadotropin beta genes in hominoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

@@

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Spalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
  \caption{Gestational periods in days and average liter sizes for studied species.}
  \label{tab:gestational_periods}
\end{table}

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \human are shown in
\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within hominoids (Supplemental \cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the \human placenta
(\cref{tab:top50_human_placenta}).
% 
Eleven \acp{PSG} are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
greatest degree of expression %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\text{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
\textit{HBA2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(\cref{tab:top50_human_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

@@

We examined the expression of five Laurasiatherians: \cow, \sheep,
\pig, \horse, and \dog. We were unable to detect any significant
differences after correcting for multiple testing between
Laurasiatherians and all other analyzed species
(Supplemental \cref{tab:laurasiatheria_lineage_specific}) or between
Laurasiatherians and Euarchontoglires
(Supplemental \cref{tab:laurasiatheria_euarchontoglires_lineage_specific}).

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Equus Caballus} are shown in
\cref{tab:top50_horse_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_equus_caballus_placenta}. The gene
with the highest expression was the relaxin,
\Sexpr{gene.expression("equus_caballus","RLN",complete=TRUE)},
followed by two miRNA
(\Sexpr{gene.expression("equus_caballus","ENSECAG00000025328",complete=TRUE)}
and
\Sexpr{gene.expression("equus_caballus","ENSECAG00000027056",complete=TRUE)})
with no known \textit{in silico} targets
\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget}, the thymosin
\Sexpr{gene.expression("equus_caballus","TMSB4",complete=TRUE)},
members of the insulin growth factor family
(\Sexpr{gene.expression("equus_caballus","IGF2",complete=TRUE)} and
\Sexpr{gene.expression("equus_caballus","IGFBP3",complete=TRUE)}) and
multiple members of annexin complexes
(\Sexpr{gene.expression("equus_caballus","S100A6",complete=TRUE)},
\Sexpr{gene.expression("equus_caballus","S100A1",complete=TRUE)}, and
\Sexpr{gene.expression("equus_caballus","ANXA2",complete=TRUE)}),
similar to the pattern of expression seen in \pig and \dog.

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently.
%
Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express 1:1
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}).

@@

@@

The \opossum placenta shows expression of genes that either do not
exist in eutherians or are not found in the term eutherian placenta
(\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
of hemoglobin in fetal tissues at term, as evidenced by the expression
of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
generally due to their very short gestation periods. Embryonic
hemoglobins in marsupials have a lower affinity for oxygen and allow
for a lower rate of oxygen uptake from maternal tissues
\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
uncharacterized gene that shows high expression (\textit{LOC10001195}) is not
found in eutherians; Ensembl gene trees reveal that homologs of this
gene exist in other amniotes such as birds and turtles, but are not
present in eutherian mammals. The homolog of this gene in chicken is
found to expressed in reproductive tract
\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
family and involved in innate immunity. It is expressed in the mammary
gland of the tammar wallaby during pregnancy and early lactation
\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and present in
relative abundance in \opossum placenta.

@@

@@

The \human placenta is distinctive from many other primates in that it
possesses a deeply invasive trophoblast that penetrates the inner
myometrium of the uterus
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
hypothesized that this feature is related to the increased demand for
resources by the developing fetus to feed the expanded size of the
human brain
\cite{Carter.ea2011:Evolutionofinvasiveplacentation,Robillard.ea2002:Evolutionaryadaptationstopre-eclampsiaeclampsia}.
\human have the largest brain in terms of relative size and this
differs markedly from closely related great apes
\cite{Boddy.ea2012:Comparativeanalysisofencephalization}. Deep
penetration of the myometrium occurs in chimpanzees and gorillas, but
this penetration may be accelerated in humans
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
can be attributed to the lack of adequate invasion of maternal tissue,
and is noted by the onset of hypertension and proteinuria
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
pre-eclampsia can develop into eclampsia, a more severe condition
resulting in seizures during pregnancy and death. Lack of invasion in
pre-eclampsia may be attributed to the inefficiency of growth
regulators as well as the inability to sufficiently suppress the
maternal immune response. Although there are speculative cases of
eclampsia in great apes
\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
it is exceedingly rare in these species
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.

@@

There is diversity in the placental expression of hormones, including
estrogen, progesterone, prolactin, relaxin, natriuretic peptides, and
the growth hormone family amount the species and lineages analyzed. We
also identified genes differentially expressed on the \human lineage
which are implicated in pre-eclampsia, including \textit{ADAM12},
\textit{CRH}, and \textit{KISS1}.

>>>>

  \parttitle{Conclusions} A diverse array of genes are differentially
  expressed and are likely to play a role in the anatomical
  variability of the placenta in the placentas of mammalian lineages.
  However, using the strictest filtering strategy the expression of only
  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
  genes is essential for the proper functioning of the placenta in all
  species analyzed.

\DLA{Include something about pre-eclampsia}

@@

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form
\cite{Mossman.1937:comparativemorphogenesisofthe,Moffett.ea2006:Immunologyofplacentationin}. Species differ in
the shape of attachment (i.e., discoid, diffuse, cotyledonary,
zonary), degree of intimacy between fetal tissues and maternal tissues
(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
interdigitation of maternal-fetal tissue (i.e., villous, folded,
trabecular, lamellar, labrynthine)
\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
mammals differ drastically in the number and size of their offspring,
as well as the length of gestation, traits often associated with
specific aspects of placental morphology
\cite{Capellini.ea2011:Placentationandmaternalinvestment}
(\cref{tab:gestational_periods}).

The placenta arose at some point before the origin of extant mammals,
a group that includes marsupials, monotremes, and eutherian or
“placental mammals”. The term “placental mammals” is misleading, as
extant monotremes (such as \textit{Ornithorhynchus anatinus}, the
platypus) possess a rudimentary placenta, as do marsupials, although
the direct apposition to maternal tissues lasts for a proportionally
shorter time during gestation
\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
Marsupials also differ from eutherian mammals in that most possess
only a choriovitelline placenta, produced from the fusion of the
chorion and yolk sac, during most of gestation. Although some
eutherian mammals may possess a choriovitelline placenta at some point
in development, the major site of nutrient exchange in placentals
occurs via the chorioallantoic placenta, produced from the fusion of
the chorion and allantois
\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
the first placenta was more like that of marsupials or eutherians;
however, using phylogenetic methods, the ancestral eutherian was
reconstructed as possessing a relatively invasive hemochorial
placenta, as in humans and mouse
\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
in \cow, genes related to the thymus-specific serine protease
\textit{PRSS16} in \textit{Loxodona africana} (in this work),
 and the expansion of the growth hormone
and chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

@@

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & \multicolumn{1}{p{1in}}{\centering Average Litter Size}  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Spalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
  \caption{Gestational periods in days and average liter sizes for studied species.}
  \label{tab:gestational_periods}
\end{table}

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \human are shown in
\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
the highest \ac{FPKM} value
%
(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
%
in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
three of which are members of a distinct expansion of the growth
hormone gene family within catarrhines (Supplemental \cref{fig:galectins_evolution_tree}). Another member of the growth
hormone gene family, \ac{GH2} (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
is also expressed to a high degree in the \human placenta
(\cref{tab:top50_human_placenta}).
% 
Eleven \acp{PSG} are highly-expressed
(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
greatest degree of expression %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
and the pseudogene \textit{PSG10P} showing the least %
(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
These
genes are members of a hominoid-specific expansion of the \ac{CEA} family
and known to be expressed by the syncytiotrophoblast
\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
%
Genes involved in the production of estrogens (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
% 
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
% 
) and progesterone (%
\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
%
show exceptionally high expression in the human placenta, as well as %
%
\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
%
which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
producing the beta subunits of the chorionic gonadotropins
(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
\textit{CGB7}, \textit{CGB8}) have reasonable expression
($\text{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
expression is on the order of the expression of
\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
%
%
Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. \textit{HBA1}, \textit{HBA2}, \textit{HBB},
and \textit{HBG2}), the maternally imprinted long noncoding RNA gene
\textit{H19}, genes coding for proteins required for growth and/or
cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
glucose transport (\textit{SLC2A1}) among others
(\cref{tab:top50_human_placenta}, Supplemental
\cref{tab:all_homo_sapiens_placenta}).

@@

We examined the expression of five laurasiatherians: \cow, \sheep,
\pig, \horse, and \dog. We were unable to detect any significant
differences after correcting for multiple testing between
laurasiatherians and all other analyzed species
(Supplemental \cref{tab:laurasiatheria_lineage_specific}) or between
laurasiatherians and Euarchontoglires
(Supplemental \cref{tab:laurasiatheria_euarchontoglires_lineage_specific}).

@@

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \textit{Equus caballus} are shown in
\cref{tab:top50_horse_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_equus_caballus_placenta}. The gene
with the highest expression was the relaxin,
\Sexpr{gene.expression("equus_caballus","RLN",complete=TRUE)},
followed by two miRNA
(\Sexpr{gene.expression("equus_caballus","ENSECAG00000025328",complete=TRUE)}
and
\Sexpr{gene.expression("equus_caballus","ENSECAG00000027056",complete=TRUE)})
with no known \textit{in silico} targets
\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget}, the thymosin
\Sexpr{gene.expression("equus_caballus","TMSB4",complete=TRUE)},
members of the insulin growth factor family
(\Sexpr{gene.expression("equus_caballus","IGF2",complete=TRUE)} and
\Sexpr{gene.expression("equus_caballus","IGFBP3",complete=TRUE)}) and
multiple members of annexin complexes
(\Sexpr{gene.expression("equus_caballus","S100A6",complete=TRUE)},
\Sexpr{gene.expression("equus_caballus","S100A1",complete=TRUE)}, and
\Sexpr{gene.expression("equus_caballus","ANXA2",complete=TRUE)}),
similar to the pattern of expression seen in \pig and \dog.

@@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all species studied and was
not a human housekeeping gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently.
%
Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express 1:1
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}).

@@

\DLA{discuss progesterone withdrawal and the lou muglin review}

@@

The \opossum placenta shows expression of genes that either do not
exist in eutherians or are not expressed in the term eutherian placenta
(\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
of hemoglobin in fetal tissues at term, as evidenced by the expression
of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
generally due to their very short gestation periods. Embryonic
hemoglobins in marsupials have a lower affinity for oxygen and allow
for a lower rate of oxygen uptake from maternal tissues
\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
uncharacterized gene that shows high expression (\textit{LOC10001195}) is not
found in eutherians; Ensembl gene trees reveal that homologs of this
gene exist in other amniotes such as birds and turtles, but are not
present in eutherian mammals. The homolog of this gene in chicken is
found to expressed in reproductive tract
\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
family and involved in innate immunity. It is expressed in the mammary
gland of the tammar wallaby during pregnancy and early lactation
\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and present in
relative abundance in \opossum placenta.

@@

\DLA{Alu was primate specific? Check to see whether this is real.}

@@

The \human placenta is distinctive from many other primates in that it
possesses a deeply invasive trophoblast that penetrates the inner
myometrium of the uterus
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
hypothesized that this feature is related to the increased demand for
resources by the developing fetus to feed the expanded size of the
human brain
\cite{Carter.ea2011:Evolutionofinvasiveplacentation,Robillard.ea2002:Evolutionaryadaptationstopre-eclampsiaeclampsia}.
\human have the largest brain in terms of relative size and this
differs markedly from closely related apes
\cite{Boddy.ea2012:Comparativeanalysisofencephalization}. Deep
penetration of the myometrium occurs in chimpanzees and gorillas, but
this penetration may be accelerated in humans
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
the United States
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
can be attributed to the lack of adequate invasion of maternal tissue,
and is noted by the onset of hypertension and proteinuria %
%
\DLA{Discuss the new ACOG requirements for diagnosis which does not
  require proteinuria.}
%
\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
pre-eclampsia can develop into eclampsia, a more severe condition
resulting in seizures during pregnancy and death. Lack of invasion in
pre-eclampsia may be attributed to the inefficiency of growth
regulators as well as the inability to sufficiently suppress the
maternal immune response. Although there are speculative cases of
eclampsia in great apes
\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
it is exceedingly rare in these species
\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.

@@

There is diversity in the placental expression of hormones, including
estrogen, progesterone, prolactin, relaxin, the growth hormone family, and
natriuretic peptides among the species and lineages analyzed.
We also identified genes differentially expressed on the \human
lineage which are implicated in pre-eclampsia, including
\textit{ADAM12}, \textit{CRH}, and \textit{KISS1}.
